---
title: "Markov Chain"
author: "Seyed Rumaiz"
output: pdf_document
date: "2026-04-19"
---

## Markov Chain Model

A discrete-time Markov chain is a stochastic process $\{X_n\}_{n \ge 0}$ that satisfies the Markov property:

$$
P(X_{n+1} = j \mid X_n = i, X_{n-1}, \dots, X_0)
= P(X_{n+1} = j \mid X_n = i)
$$

This means that the future state depends only on the current state.

## Transition Matrix

The system is described by a transition probability matrix $P$:

$$
P =
\begin{bmatrix}
p_{11} & p_{12} & \cdots & p_{1k} \\
p_{21} & p_{22} & \cdots & p_{2k} \\
\vdots & \vdots & \ddots & \vdots \\
p_{k1} & p_{k2} & \cdots & p_{kk}
\end{bmatrix}
$$

where

$$
p_{ij} = P(X_{n+1} = j \mid X_n = i), \quad \sum_{j=1}^{k} p_{ij} = 1
$$

## State Distribution

Let the initial distribution be:

$$
\pi^{(0)} = (\pi_1, \pi_2, \dots, \pi_k)
$$

Then the distribution after one step is:

$$
\pi^{(1)} = \pi^{(0)} P
$$

After $n$ steps:

$$
\pi^{(n)} = \pi^{(0)} P^n
$$

## Steady-State Distribution

The steady-state (stationary) distribution $\pi$ satisfies:

$$
\pi P = \pi
$$

with the constraint:

$$
\sum_{i=1}^{k} \pi_i = 1
$$

This is the long-run behaviour of the Markov chain.

## Implementation

Lets implement the Markov Chain with some examples!
We will use the `markovchain` package, along with the `expm` package for matrix power.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install packages
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("markovchain")
install.packages("expm")
```

```{r}
# Load libraries
library(markovchain)
library(expm)
```

```{r}
# Define transition matrix
states <- c("Sunny","Cloudy","Rainy")

transitionMatrix <- matrix(c(
  0.7,0.2,0.1,
  0.3,0.4,0.3,
  0.2,0.3,0.5),
  nrow=3, byrow=TRUE)

transitionMatrix
```

```{r}
# Create a Markov Chain object
weatherMC <- new("markovchain",
                 states = states,
                 transitionMatrix = transitionMatrix,
                 name = "Weather")

weatherMC
```

```{r}
# Predict next state probabilities
initialState <- c(1,0,0)  # Sunny

nextDay <- initialState %*% transitionMatrix
nextDay
```

```{r}
# Weather after 3 days
transitionMatrix3 <- transitionMatrix %^% 3
transitionMatrix3
```

```{r}
# Long run probabilities
steadyStates(weatherMC)
```

```{r}
# Simulate a Markov Chain sequence (generate weather for 10 days)
set.seed(123)
weatherSimulation <- rmarkovchain(n=10, object=weatherMC, t0="Sunny")
weatherSimulation
```

```{r}
# Plot the Markov Chain
plot(weatherMC)
```

```{r}
# Find the probability after 2 steps starting from Rainy
initial <- c(0,0,1)   # Rainy

P2 <- transitionMatrix %^% 2

result <- initial %*% P2
result
```
